Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-02-03 analysis/paper/paper.Rmd
worked on the intro section

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange or introduction of goods, species, and ideas that could result in far-reaching impacts on local societies economically and culturally. In Southwest China, the Hengduan Mountain ranges as natural passes play an important role in the migration of farming communities and the spread of crops into this region and mainland Southeast Asia. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022]. However, how and to what extent the two crops were incorporated into the local cultures is still not fully understood. To explore whether the new species became main food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, the oldest Neolithic site in the Yanyuan region so far excavated.

<!-- sites and research questions-->
As an early Neolithic site, Guijiabao (ca. 4800-4300 BP) is crucial to understanding cultural interactions and the spread of crops into the Yanyuan Basin due to its location at the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau. The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@Guijiabao2016], different from the abundant rice grains found at other Neolithic sites in that area. Although the plant remains suggest the possible domestication of millet in the early Neolithic, we still lack direct evidence of culinary practices during the Neolithic transition. To explore the extent of acceptance of new crops and species in the Yanyuan Basin, we ask the questions: What was the culinary practices during the Neolithic transition? Could we identify the evidence of new crops or species incorporated into the local dietary systems in the Early Neolithic period? We answer those questions by examining potsherds collected from Guijiabao using organic residue analysis, which could offer a better understanding of food sources during the Neolithic transition in the Yanyuan Basin.

This study aims to detect and identify organic residues, particularly lipid residues, that could have been preserved in the ceramic fabric to explore the vessel usage at Guijiabao. Lipid analysis is useful in archaeology since lipid molecules can be well preserved over thousand years in the porous fabric of pottery with the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. Lipid residues are the most direct evidence to investigate culinary practices and dietary patterns since specific lipid compounds can be associated with different types of foodstuffs cooked or stored in pottery vessels [@Copley2005]. It is especially informative for human subsistence in the situation where archaeological sites lack organic remains due to acidic sediments [@Kwak2020]. The growing application of organic residue analysis demonstrate it offers a better understanding of culinary practices during the transitional period to agricultural societies [@Craig2011], dietary patterns in

@@

<!--Yanyuan area-->
The Yanyuan area in Sichuan Province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)). Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022]. Therefore, the archaeological sites in the Yanyuan Basin are important to understanding the cultural interactions between southwest China and other regions, including the Sichuan basin, northwest China, and northwest Yunnan.

<!--Guijiabao site-->
Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is a unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long-term occupation [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (ca. 5000-3700 BP), the Bronze Age (ca. 3200-2700 BP), and the late North Song to early Ming Dynasties (ca. 1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further separated into two subphases, 5000-4500 ca. BP and 4500-3700 ca. BP, according to the content of artifact assemblages and ceramic forms and styles.

<!--the major finding at GJB-->
Among the various ceramic styles, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a]. The archaeobotanical remains reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016]. The adoption of millet cultivation can be traced back to ca. 5000 BP [@Hao2022]. In addition, rice grains (*Oryza sativa*) have also been found in ash pits throughout different phases where the earliest rice dates to 4700 ca. BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to Neolithic period (5000 ca. BP to 3700 ca. BP) further suggests the presence of rice phytoliths throughout the whole period in addition to millet phytoliths, indicative of a mixed millet/rice cultivation prior to 5000 ca. BP than previously thought [@Huan2022].

@@

Despite of a long-standing debate between waves of rice/millet dispersal from northwest China and from the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP. and to western Yunnan before 4600 BP. [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010]. The current finding at Guijiabao and Gaoshan support this argument. However, plant remains from Guijiabao shows that millet is the major crop, which is at odds with the predominance of rice based on phytolich analysis and other macro-botanical analyses from Gaoshan or sites in Sichuan and Yunnan, such as Baodun, Henglanshan and Baiyangcun [@Hao2022; @Guedes2013]. Although the analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, evidence directly reflects food preparation or processing using ceramics still lacks. This study aims to provide direct evidence from food residues in ceramics to investigate food choices associated with new practices of crop cultivation and local subsistence.

@@

The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them (Table \@ref(tab:FA-composition-table)). The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds in most potsherds (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, some potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. A larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005]. Although oleic acid (C18:1) is the most abundant compound for most of our samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.

@@

@@

Long-chain fatty acids from C20:0 to C32:0 are present in a pottery sample (SYG-TN13-E22-2#2) with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could be also associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998; @Suryanarayan2021]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.

@@

<!--discussion about possible plant origins-->
One aim of our study is to explore whether new crops, such as C3 plants (eg. rice), C4 plants (eg. millets), or new species (eg. pigs/horses) became an essential part of local diet in Yanyuan during the Neolithic transition. We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies [@Shoda2018; @Kwak2020]. It is also possible other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016]. Surprisingly, we only observed signals of C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022]. Due to our small samples, it is hard to determine the role of rice/millet during the Neolithic period. Although freshwater fish could be another origin of non-ruminant fats that can suggested by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Kwak2017].

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. But whether horse was part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress that can provide further information.

@@

@@

Our study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in a relatively low amount for most potsherds. The food cooked in those ceramics include ruminant fats and terrestrial plants during the Neolithic transition when the mixing farming of rice/millet spread into southwest China. It is most likely the terrestrial plant signal is C3 plant, implying rice processing in the pot. However, the evidence of direct processing of terrestrial plants is not as much as we expected according to the large quantity of plant remains, such as millet grains, found at Guijuabao. Instead, food processing in the early Neolithic is mostly ruminant fats that could be from local wild animals, such as deer. This might indicate that the proportion of rice/millet in the local diet is not high during the Neolithic transition, suggesting that hunting might be the primary subsistence, supplemented by rice/millet farming. This study offers a perspective from organic residue analysis to explore food choices and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period. We recognize the limitation of our interpretation of the local diet due to the small sample size. A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.

>>>>

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could result in far-reaching economic or cultural impacts on local societies. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP. and to western Yunnan before 4600 BP. [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022]. Therefore, the Neolithic Hengduan Mountain areas are key to understanding the processes of agriculture expansion. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022]. However, how and to what extent the two crops were incorporated into the local cultures is still not fully understood.

To explore whether the new crops or species became important food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@Guijiabao2016], different from the abundant rice grains found at other Neolithic sites in that area. Although the analysis of plant remains suggests millet/rice farming in the early Neolithic, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food choices in the Early Neolithic period? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels during the Neolithic transition in the Yanyuan Basin.

The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies or culinary practices across a region [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers to trace their original foodstuffs [@Copley2005; @Whelton2021]. This is because lipid molecules can be accumulated and well preserved in the porous fabric of pottery over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detect and identify lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020].

@@

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao site located in the Yanyuan Basin in Sichuan Province. Yanyuan Basin is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)). River valleys of Hengduan Mountain provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022]. Among the archaeological sites in Yanyuan, Guijiabao (ca. 4800-4300 BP) located in the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau is a unique site with continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site with long-term occupation [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (ca. 5000-3700 BP), the Bronze Age (ca. 3200-2700 BP), and the late North Song to early Ming Dynasties (ca. 1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further separated into two subphases, 5000-4500 ca. BP and 4500-3700 ca. BP, according to the content of artifact assemblages and ceramic forms and styles.

<!--the major finding at GJB-->
At Guijiabao, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a]. The archaeobotanical remains reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016]. The adoption of millet cultivation can be traced back to ca. 5000 BP [@Hao2022]. In addition, rice grains (*Oryza sativa*) have also been found in ash pits throughout different phases where the earliest rice dates to 4700 ca. BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to Neolithic period (5000 ca. BP to 3700 ca. BP) further suggests the presence of rice phytoliths throughout the whole period in addition to millet phytoliths, indicative of a mixed millet/rice cultivation prior to 5000 ca. BP than previously thought [@Huan2022].

@@

 The current finding at Guijiabao and Gaoshan support this argument. However, plant remains from Guijiabao shows that millet is the major crop, which is at odds with the predominance of rice based on phytolich analysis and other macro-botanical analyses from Gaoshan or sites in Sichuan and Yunnan, such as Baodun, Henglanshan and Baiyangcun [@Hao2022; @Guedes2013]. Although the analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, evidence directly reflects food preparation or processing using ceramics still lacks. This study aims to provide direct evidence from food residues in ceramics to investigate food choices associated with new practices of crop cultivation and local subsistence.

@@

The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them (Table \@ref(tab:FA-composition-table)). The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds in most potsherds (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, a number of potsherds presents higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. Although a larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005], it is hard to directly relate the pattern we observed to a plant origin based on GC-MS results alone.

@@

quant_FA_number <-
  quant_FA %>%
  pivot_longer(!number&!sample,
               names_to = "FA", 
               values_to = "weight") %>% 
  select(-number) %>% 
  filter(str_starts(FA, "ng") & !sample =="") %>% 
  mutate(FA = str_remove(FA, "ng "),
         weight = round(weight/1000, 2)) %>% 
  pivot_wider(names_from = sample, 
              values_from = weight)

FA_max_sum <-
  sum(quant_FA_number$"SYG-TN13-E22-2#2")

@@

One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of samples, suggesting a good preservation of lipids. Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could be also associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. This also corresponds to the lipid distribution of the same potshard which reflects a possible plant origin based on the GC-MS results. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.

@@

<!--discussion about possible plant origins-->
One aim of our study is to explore whether new crops, such as C3 plants (eg. rice), C4 plants (eg. millets), or new species (eg. pigs/horses) became an essential part of local diet in Yanyuan during the Neolithic transition. We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies [@Shoda2018; @Kwak2020]. It is also possible other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016]. Surprisingly, we only observed signals of C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022]. Due to our small samples, it is hard to determine the role of rice/millet during the Neolithic period. Although freshwater fish could be another origin of non-ruminant fats that can suggested by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. But whether horse was part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on C3 diet based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress that can provide further information.

@@

-adoption of farming in a relatively enclosed circumstances?

@@

Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in low concentrations for most potsherds. The food cooked in those ceramics include ruminant fats and terrestrial plants during the Neolithic transition when the mixing farming of rice/millet spread into southwest China. It is most likely the terrestrial plant signal is C3 plant, implying rice processing in the pot. However, the evidence of direct processing of terrestrial plants is not as much as we expected according to a relatively large quantity of plant remains found at Guijiabao, such as millet grains. Instead, food processing in the early Neolithic is mostly ruminant fats that could originate from local wild animals, such as deer. This indicates that the proportion of rice/millet in the local diet is not high during the Neolithic transition, suggesting that hunting might be the primary subsistence, supplemented by rice/millet farming. We recognize the limitation of our interpretation of the local diet due to the small sample size. However, our study can still offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period. A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.
